\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{isabelle,isabellesym}
\usepackage{amsmath}
\usepackage{amsthm}
\newcommand{\size}[1]{\lvert#1\rvert}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\expectation}{\mathrm{E}}

% further packages required for unusual symbols (see also
% isabellesym.sty), use only when needed

%\usepackage{amssymb}
  %for \<leadsto>, \<box>, \<diamond>, \<sqsupset>, \<mho>, \<Join>,
  %\<lhd>, \<lesssim>, \<greatersim>, \<lessapprox>, \<greaterapprox>,
  %\<triangleq>, \<yen>, \<lozenge>

%\usepackage{eurosym}
  %for \<euro>

%\usepackage[only,bigsqcap,fatsemi,interleave,sslash]{stmaryrd}
  %for \<Sqinter>, \<Zsemi>

%\usepackage{eufrak}
  %for \<AA> ... \<ZZ>, \<aa> ... \<zz> (also included in amssymb)

%\usepackage{textcomp}
  %for \<onequarter>, \<onehalf>, \<threequarters>, \<degree>, \<cent>,
  %\<currency>

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

% for uniform font size
%\renewcommand{\isastyle}{\isastyleminor}


\begin{document}

\title{Formalization of Randomized Approximation Algorithms for Frequency Moments}
\author{Emin Karayel}
\begin{abstract}
In 1999 Alon et.\ al.\ introduced the still active research topic of approximating the frequency moments of a data stream using randomized algorithms with minimal spage usage.
This includes the problem of estimating the cardinality of the stream elements---the zeroth frequency moment.
But, also higher order frequency moments that provide information about the skew of the data stream, which is for example critical information for parallel processing.
The frequency moment of a data stream $a_1, \ldots, a_m \in U$ can be defined as $F_k := \sum_{u \in U} C(u,a)^k$ where $C(u,a)$ is the count of occurences of $u$ in the stream $a$.
They introduce both lower bounds and upper bounds, which were later improved by newer publications.
The algorithms have guaranteed success probability and accuracy, without making any assumptions on the input distribution.
They are an interesting use-case for formal verification, because they rely on deep results from both algebra and analysis, require a large body of existing results.
This work contains the formal verification of three algorithms for the approximation of $F_0$, $F_2$ and $F_k$ for $k \geq 3$.
To achieve it, the formalization also includes reusable components common to all algorithms, such as universal hash families, the median method, formal modelling of one-pass data stream algorithms and a generic flexible encoding library for the verification of space complexities.
\end{abstract}

\maketitle

\tableofcontents

% sane default for proof documents
\parindent 0pt\parskip 0.5ex

% generated text of all theories
\input{session}
\appendix
\section{Informal proof of correctness for the $F_0$ algorithm\label{sec:f0_proof}}
This section contains a detailed informal proof for the correctness of the $F_0$-algorithm.
Because of the standard amplification result about medians (see for example \cite{alon1999}) it
is enough to show that each of the estimates the median is taken from is within the desired interval
with success probability $\frac{2}{3}$.

To verify the latter, let $a_1, \ldots, a_m$ be the stream elements, where we assume that the
elements are a subset of $\{0,\ldots,n-1\}$ and $0 < \delta < 1$ be the desired relative accuracy.
Let $p$ be the smallest prime such that $p \geq \max (n,19)$ and let $h$ be a random polynomial over
$GF(p)$ with degree strictly less than $2$.
The algoritm also introduces the internal parameters $t, r$ defined by:
\begin{eqnarray*}
    t & := & \lceil 80\delta^{-2} \rceil \\
    r & := & 4 \log_2 \lceil \delta^{-1} \rceil + 24
\end{eqnarray*}
The estimate the algorithm obtains is:
\begin{align*}
    A := & \left\{ a_1, \ldots, a_m \right\} &
    H := & \left\{ \lfloor h(a) \rfloor_r \middle \vert a \in A \right\} \\
    R := & \begin{cases} t p \left(\mathrm{min}_t (H) \right)^{-1} & \textrm{ if } \size{H} \geq t \\
    \size{H} & \textrm{ othewise,} \end{cases} &
\end{align*}
Here $\mathrm{min}_t(H)$ denotes the $t$-th smallest element of $H$.
With these definitions, it is possible to state the goal as:
\[
    P(\size{R - F_0} \leq \delta \size{F_0}) \geq \frac{2}{3} \textrm{.}  
\]
which is shown by separately in the following two subsections for the cases $F_0 \geq t$ and $F_0 < t$.
\subsection{Case $F_0 \geq t$}
Let us introduce:
\begin{eqnarray*}
    H^* & := & \left\{ h(a) \middle \vert a \in A \right\}^{\#} \\
    R^* & := & tp \left( \mathrm{rank}^{\#}_t(H^*) \right)^{-1}
\end{eqnarray*}
These definitions correspond to the $H$, $R$ but with a few minor modifications.
The set $H^*$ is a multiset, this means that each element also has a multiplicity, counting the
number of \emph{distinct} elements of $A$ being mapped by $h$ to the same value.
Note that by definition: $\size{H^*}=\size{A}$.
Similarly the operation $\mathrm{min}^{\#}_t$ obtains the $t$-th element of the multiset $H$
(taking multiplicities into account).
Note also that there is no rounding operation $\lfloor \cdot \rfloor_r$ in the definition of $H^*$.
The key reason for the introduction of these alternative versions of $H, R$ is that it is easier to
show probabilistic bounds on the distances $\size{R^* - F_0}$ and $\size{R^* - R}$ as opposed to 
$\size{R - F_0}$ directly.
In particular the plan is to show:
\begin{eqnarray}
 \delta' & := & \frac{3}{4}\delta \\
 P\left(\size{R^*-F_0} > \delta' F_0\right) & \leq & \frac{2}{9} \textrm{, and} \label{eq:r_star_dev} \\
 P\left(\size{R^*-F_0} \leq \delta' F_0 \wedge \size{R-R^*} > \frac{\delta}{4} F_0\right) & \leq & \frac{1}{9} \label{eq:r_star_r}
\end{eqnarray}
I.e. the probability that $R^*$ has not the relative accuracy of $\frac{3}{4}\delta$ is less that $\frac{2}{9}$ and the probability 
that assuming $R^*$ has the relative accuracy of $\frac{3}{4}\delta$ but that $R$ deviates by more that $\frac{1}{4}\delta F_0$ is at most $\frac{1}{9}$.
Hence, the probability that neither of these events happen is at least $\frac{2}{3}$ but in that case:
\begin{equation}
    \label{eq:concl}
    \size{R-F_0} \leq \size{R - R^*} + \size{R^*-F_0} \leq \frac{\delta}{4} F_0 + \frac{3 \delta}{4} F_0 = \delta F_0 \textrm{.}
\end{equation}

For the verification of \autoref{eq:r_star_dev} let us introduce:
\[
    Q(u) = \size{\left\{h(a) < u \mid a \in A \right\}}
\]
and observe that $\mathrm{min}_t^{\#}(H^*) < u$ if $Q(u) \geq t$ and $\mathrm{min}_t^{\#}(H^*) \geq v$ if $Q(v) \leq t-1$.
To see why this is true note that, if at least $t$ elements of $A$ are mapped by $h$ below a certain value, then the rank $t$ element must also be within them, and thus also be below that value.
And that the opposite direction of this conclusion is also true.
Note that this relies on the fact that $H^*$ is a multiset and that multiplicities are being taken into account, when computing the $t$-th smallest element. 

Alternatively, it is also possible to write $Q(u) = \sum_{a \in A} 1_{\{h(a) < u\}}$\footnote{The notation $1_A$ is shorthand for the indicator function of $A$, i.e., $1_A(x)=1$ if $x \in A$ and $0$ otherwise.}, i.e., $Q$ is a sum of pairwise independent $\{0,1\}$-valued random variables, with expectation $\frac{u}{p}$ and variance $\frac{u}{p} - \frac{u^2}{p^2}$.
\footnote{A consequence of $h$ being choosen uniformly from a $2$-independent hash family.}
Using lineariy of expectation and Bienaym\'e's identity, it follows that $\var \, Q(u) \leq \expectation \, Q(u) = |A|u p^{-1} = F_0 u p^{-1}$ for $u \in \{0,\ldots,p\}$.

For $v = \left\lfloor \frac{tp}{(1-\delta') F_0} \right\rfloor$ it is possible to conclude:
\begin{eqnarray*}
    t-1 & \leq\footnotemark & \frac{t}{(1-\delta')} - 3\sqrt{\frac{t}{(1-\delta')}} - 1 \\
     &\leq&  \frac{F_0 v}{p} - 3 \sqrt{\frac{F_0 v}{p}} \leq \expectation Q(v) - 3 \sqrt{\var Q(v)}
\end{eqnarray*}
\footnotetext{The verification of this inequality is a lengthy but straightforward calculcation using the definition of $\delta'$ and $t$.}
and thus using Tchebyshev's inequality:
\begin{align}
    P\left(R^* < \left(1-\delta'\right) F_0\right) & = P\left(\mathrm{rank}_t^{\#}(H^*) > \frac{tp}{(1-\delta')F_0}\right) \nonumber \\ 
    & \leq P(\mathrm{rank}_t^{\#}(H^*) \geq v) = P(Q(v) \leq t-1) \label{eq:r_star_upper_bound} \\
    & \leq P\left(Q(v) \leq \expectation Q(v) - 3 \sqrt{\var Q(v)}\right) \leq \frac{1}{9} \textrm{.} \nonumber
\end{align}
Similarly for $u = \left\lceil \frac{tp}{(1+\delta') F_0} \right\rceil$ it is possible to conclude:
\begin{eqnarray*}
    t & \geq & \frac{t}{(1+\delta')} + 3\sqrt{\frac{t}{(1+\delta')}+1} + 1 \\
     &\geq&  \frac{F_0 u}{p} + 3 \sqrt{\frac{F_0 u}{p}} \geq \expectation Q(u) + 3 \sqrt{\var Q(v)}
\end{eqnarray*}
and thus using Tchebyshev's inequality:
\begin{align}
    P\left(R^* > \left(1+\delta'\right) F_0\right) & = P\left(\mathrm{rank}_t^{\#}(H^*) < \frac{tp}{(1+\delta')F_0}\right) \nonumber \\ 
    & \leq P(\mathrm{rank}_t^{\#}(H^*) < u) = P(Q(u) \geq t) \label{eq:r_star_lower_bound} \\
    & \leq P\left(Q(u) \geq \expectation Q(u) + 3 \sqrt{\var Q(u)}\right) \leq \frac{1}{9} \textrm{.} \nonumber
\end{align}
To verfiy \autoref{eq:r_star_r}, note that
\begin{equation}
    \label{eq:rank_eq}
    \mathrm{min}_t(H) = \lfloor \mathrm{min}_t^{\#}(H^*) \rfloor_r
\end{equation}
if there are no collisions, induced by the application of $\lfloor h(\cdot) \rfloor_r$ on the elements of $A$.
Even more carefully, note that the equation would remain true, as long as there are no collision within the smallest $t$ elements of $H^*$.
Because \autoref{eq:r_star_r} needs to be shown only in the case where $R^* \geq (1-\delta') F_0$, i.e., when $\mathrm{min}_t^{\#}(H^*) \leq v$,
it is enough to bound the probability of a collision in the range $[0; v]$.
Moreover \autoref{eq:rank_eq} implies $\size{\mathrm{min}_t(H) - \mathrm{min}_t^{\#}(H^*)} \leq \max(\mathrm{min}_t^{\#}(H^*), \mathrm{min}_t(H)) 2^{-r}$ from
which it is possible to derive $\size{R^*-R} \leq \frac{\delta}{4} F_0$.
% R* = tp/rank_t#, R = tp/rank_t => |R-R*| = | tp [ rank_t - rank_t# / rank_t rank_t# ] | <= 
% tp \max{rank_t, rank_t#} 2^-r / rank_t rank_t# <= t 2^-r <= t 2^-24 d'^4 <= F_0 d'^4 2^-24 <= 1/4 d' F_0
%
%Let's summarize: If
%\begin{eqnarray*}
%    R^* & \geq & (1-\delta') F_0 \\
%    \lfloor h(a) \rfloor_r & \neq & \lfloor h(b) \rfloor_r \textrm{ for } a \neq b \in A \wedge h(a) \leq v \wedge h(b) \leq v
%\end{eqnarray*}
%then $\mathrm{rank}_t(H) = \lfloor \mathrm{rank}_t^{\#}(H^*) \rfloor_r$.
Another important fact is that $h$ is injective with probability $1-\frac{1}{p}$, this is because $h$ is choosen uniformly from the polynomials of degree less than $2$.
If it is a degree $1$ polynomial, it is a linear function on $GF(p)$ and thus injective.
Because $p \geq 18$ the probability that $h$ is not injective can be bounded by $1/18$.
However, even if $h$ is injective, there is still a possibility of collision, because of the application of the rounding operation $\lfloor \cdot \rfloor_r$.
The plan is to bound that probability by $1/18$ as well to show \autoref{eq:r_star_r}.
\begin{eqnarray*}
    P\left( \size{R^*-F_0} \leq \delta' F_0 \wedge \size{R-R^*} > \frac{\delta}{4} F_0 \right) & \leq & \\
    P\left( R^* \geq (1-\delta') F_0 \wedge \mathrm{min}_t^{\#}(H^*) \neq \mathrm{min}_t(H) \wedge h \textrm{ inj.}\right) + P(\neg h \textrm{ inj.}) & \leq & \\
    P\left( \exists a \neq b \in A. \lfloor h(a) \rfloor_r = \lfloor h(b) \rfloor_r \leq v \wedge h(a) \neq h(b) \right) + \frac{1}{18} & \leq & \\
    \frac{1}{18} + \sum_{a \neq b \in A} P\left(\lfloor h(a) \rfloor_r = \lfloor h(b) \rfloor_r \leq v \wedge h(a) \neq h(b) \right) & \leq & \\
    \frac{1}{18} + \sum_{a \neq b \in A} P\left(\size{h(a) - h(b)} \leq v 2^{-r} \wedge h(a) \leq v (1+2^{-r}) \wedge h(a) \neq h(b) \right) & \leq & \\
    \frac{1}{18} + \sum_{a \neq b \in A} \sum_{\substack{a', b' \in \{0,\ldots, p-1\} \wedge a' \neq b' \\ \size{a'-b'} \leq v 2^{-r} \wedge a' \leq v (1+2^{-r})}} P(h(a) = a') P(h(b)= b') & \leq & \\
    \frac{1}{18} + 6 \frac{F_0^2 v^2}{p^2} 2^{-r} & \leq & \frac{1}{9} \textrm{.}
%    96 t^2 2^{-r} + \frac{1}{18} & \leq & \frac{1}{9}
\end{eqnarray*}
Which shows that \autoref{eq:r_star_r} is true and \autoref{eq:r_star_upper_bound} and~\ref{eq:r_star_lower_bound} implies 
\autoref{eq:r_star_dev}, which means the reasoning in \autoref{eq:concl} confirms:
\begin{equation}
    P(\size{R - F_0} \leq \delta \size{F_0}) \geq \frac{2}{3}
\end{equation}

The following subsection confirms that this is also true for the remaining case, if $F_0 < t$, concluding the proof.
\subsection{Case $F_0 < t$}
Note that in this case $\size{H} \leq F_0 < t$ and thus $R = \size{H}$, hence the goal is to show that:
$P(\size{H} \neq F_0) \leq \frac{1}{3}$.

The latter can only happen, if there is a collision induced by the application of $\lfloor h(\cdot)\rfloor_r$. As before $h$ is not injective with probability at least $\frac{1}{18}$, hence:
\begin{eqnarray*}
    P\left( \size{R - F_0} > \delta F_0\right) & \leq & \\
    P\left( R \neq F_0 \right) & \leq & \\
    \frac{1}{18} + P\left( R \neq F_0 \wedge h \textrm{ injective} \right) & \leq & \\
    \frac{1}{18} + P\left( \exists a \neq b \in A. \lfloor h(a) \rfloor_r = \lfloor h(b) \rfloor_r  \right) & \leq & \\
    \frac{1}{18} + \sum_{a \neq b \in A} P\left(\lfloor h(a) \rfloor_r = \lfloor h(b) \rfloor_r \wedge h(a) \neq h(b) \right) & \leq & \\
    \frac{1}{18} + \sum_{a \neq b \in A} P\left(\size{h(a) - h(b)} \leq p 2^{-r} \wedge h(a) \neq h(b) \right) & \leq & \\
    \frac{1}{18} + \sum_{a \neq b \in A} \sum_{\substack{a', b' \in \{0,\ldots, p-1\} \\  a' \neq b' \wedge \size{a'-b'} \leq p 2^{-r}}} P(h(a) = a') P(h(b)= b') & \leq & \\
    \frac{1}{18} + F_0^2 2^{-r+1} \leq \frac{1}{9} \textrm{.}
\end{eqnarray*}
Which concludes the proof. \qed
\bibliographystyle{abbrv}
\bibliography{root}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
